---
title: "Maoz_Week1_BST209"
output: html_document
date: "2025-07-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r merge datasets}
rm(list=ls())
library(dplyr)
load("covariate_df.rData")
d <- covariate_df
length(unique(d$SEQN)) == dim(d)[1] ##quick and dirty check that there is one row per ID
q <- read.csv("questionnaire.csv")

d <- d %>%
  left_join(select(q, SEQN, MCQ220), by = "SEQN")

table(d$MCQ220) ## check which values
d$MCQ220[d$MCQ220 == 2] <- 0  ## change unaffected to 0 to 
d$MCQ220 <- factor(d$MCQ220) ## May not need this but makes cancer a factor
d <- subset(d, is.na(MCQ220) == FALSE) ## Excludes individuals with missing data for the outcome

sum(complete.cases(d)) ## Check how many entries have all the data (n=0)

d <- d[, colSums(is.na(d)) <= 1000] ## Removes columns with a lot of missing data

d <- filter(d, complete.cases(d)) ## After removing the columns causing a lot of missing-ness, look for complete cases







library(dplyr)
library(tableone)

# Rename columns in d
d <- d %>%
  rename(
    Gender = RIAGENDR,
    Age_years = RIDAGEYR,
    Race_Ethnicity = RIDRETH1,
    Education_Level = DMDEDUC2,
    Marital_Status = DMDMARTL,
    Income_to_Poverty_Ratio = INDFMPIR,
    Household_Size = DMDHHSIZ,
    BMI_kgm2 = BMXBMI,
    Weight_kg = BMXWT,
    Height_cm = BMXHT,
    Waist_Circumference_cm = BMXWAIST,
    Systolic_BP_mmHg = BPXSY1,
    Diastolic_BP_mmHg = BPXDI1,
    Little_Interest_Pleasure = PHQ020,
    Feeling_Down_Depressed = PHQ030,
    Alcohol_Intake_g = DR1TALCO,
    Total_Cholesterol_mgdl = LBXTC,
    HDL_Cholesterol_mgdl = LBDHDD,
    Serum_Creatinine_mgdl = LBXSCR
  ) %>%
  mutate(MCQ220 = factor(MCQ220, levels = c(0, 1), labels = c("No", "Yes")))

# Define variables and factor variables with new names
myVars <- c(
  "Gender", "Age_years", "Race_Ethnicity", "Education_Level", "Marital_Status", 
  "Income_to_Poverty_Ratio", "Household_Size", "BMI_kgm2", "Weight_kg", "Height_cm", 
  "Waist_Circumference_cm", "Systolic_BP_mmHg", "Diastolic_BP_mmHg", 
  "Little_Interest_Pleasure", "Feeling_Down_Depressed", "Alcohol_Intake_g", 
  "Total_Cholesterol_mgdl", "HDL_Cholesterol_mgdl", "Serum_Creatinine_mgdl" 
)

factorVars <- c(
  "Gender", "Race_Ethnicity", "Education_Level", "Marital_Status",
  "Little_Interest_Pleasure", "Feeling_Down_Depressed"
)


## Visualizing the data
skimr::skim(d) ##Gives an overview of the data - takes a long time to run but I can see that many columns have a lot of missing values



## Making a table 1 

# Create Table 1 stratified by MCQ220 (cancer history)
tbl1 <- CreateTableOne(vars = myVars,
                       factorVars = factorVars, strata = "MCQ220",
                       data = d,
                       addOverall = TRUE)

tbl1 <- as.data.frame(print(tbl1, printToggle = FALSE, noSpaces = TRUE, smd = TRUE))




View(tbl1)
##write.xlsx(tbl1, "tbl1.xlsx", rowNames = T)


 





set.seed(1)
split <- sample(c(1,0), nrow(d), replace = TRUE, prob = c(0.7, 0.3))
check <- table(split)
check
prop.table(check)
rm(check)



```

#_______________________________________________________________________________
#Correlation Plot
#_______________________________________________________________________________

## Make a correlation plot for variables in myVars
library(corrplot)
d_corr_plot = d[, myVars]
corr_val <- cor(d_corr_plot)
corrplot(corr_val,type = "lower",order="hclust")

#_______________________________________________________________________________
#Lillian Section
#_______________________________________________________________________________

#_______________________________________________________________________________
#Table 1
#_______________________________________________________________________________


#Change Categorical and Binary Variables to Factor          
d$Gender <- factor(d$Gender, labels = c("Male", "Female"))
d$Race_Ethnicity <- factor(d$Race_Ethnicity)
d$Education_Level <- factor(d$Education_Level) 
d$Marital_Status <- factor(d$Marital_Status)
d$Household_Size <- factor(d$Household_Size)

#Render Categorical vs Continuous Variables
render.categorical <- function(x, ...) {
  c("", sapply(stats.apply.rounding(stats.default(x)), function(y) with(y,
                  sprintf("%s (%s%%)", prettyNum(FREQ, big.mark=","), PCT))))}

my.render.cont <- function(x) {
  with(stats.apply.rounding(stats.default(x, ), digits = 3),
       c("Age_years",
         "Median [IQR]" =
           sprintf(paste("%s [%s, %s]"), MEDIAN,Q1,Q3))) }

render.strat <- function (label, n, ...) {
  sprintf("<span class='stratlabel'>%s<br><span class='stratn'>(N=%s)</span></span>", 
          label, prettyNum(n, big.mark=",")) }
          


#Render Continuous Variables vs Categorical
my_render <- function(x, name, ...) {
  # Use mean ± SD for Age only
  if (name == "Age_years") {
    with(stats.apply.rounding(stats.default(x), digits = 1),
         sprintf("%s ± %s", MEAN, SD))
  } else {
    # Default: counts and percentages
    render.default(x, name, ...)
  }
}

#Rename Data Variables
library(Hmisc)
label(d$Age_years) <- "Age (years)"
label(d$Gender) <- "Gender"
label(d$Race_Ethnicity) <- "Race/Ethnicity"
label(d$Education_Level) <- "Education Level"
label(d$Marital_Status) <- "Marital Status"
label(d$Household_Size) <- "Household Size"
label(d$Income_to_Poverty_Ratio) <- "Income to Poverty"


#Table 1 Summary Table by Cancer
table1(~ Age_years  + Gender  + Race_Ethnicity + Education_Level + Marital_Status + Household_Size + Income_to_Poverty_Ratio | MCQ220, data = d, overall = "Total", caption = "Table 1: Demographic Characteristics by Cancer Status", render.var = my_render)

#_______________________________________________________________________________
#Visualizations
#_______________________________________________________________________________


#Plot of Cancer Status
ggplot(d, aes(x=as.factor(MCQ220), fill=as.factor(MCQ220) )) +
labs(
x = "Cancer Status",
y = "Count",
title = "Cancer Status"
)+
  geom_bar( ) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  scale_fill_manual(
    values = c("No" = "#FFA500", "Yes" = "#FFDB58"),  
    labels = c("Yes" = "Cancer", "No" = "No Cancer")
  ) +
  scale_x_discrete(
  labels = c("Yes" = "Cancer", "No" = "No Cancer")
)+
  theme(plot.title = element_text(hjust = 0.5), legend.position="none", panel.background = element_rect(fill = "white"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())
  


#Histogram of Cancer vs No Cancer Groups
library(ggplot2)
ggplot(df_comp, aes(x = factor(Cancer), y = Age, fill = factor(Cancer))) +
  geom_col(width = 0.6, alpha = 0.8) +  
  labs(
    x = "Cancer Status",  # X axis label
    y = "Age",  # Y axis label
    title = "Age by Cancer Status",  # Title
    fill = "Cancer Status"  # Legend title
  ) +
  scale_fill_manual(
    values = c("1" = "#FF69B4", "2" = "#1E90FF"),  # Custom colors for bar graph
    labels = c("1" = "Cancer", "2" = "No Cancer")  # Full labels for the legend
  ) +
  theme_minimal(base_size = 10) +  
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold", color = "#333333"),  
    axis.title.x = element_text(size = 12, face = "bold", color = "#333333"),  # Styled x axis title
    axis.title.y = element_text(size = 12, face = "bold", color = "#333333"),  # Styled y axis title
    axis.text = element_text(size = 10, color = "#333333"),  # Styled axis text
    legend.position = "bottom",  # Move legend 
    legend.title = element_text(size = 10, face = "bold"),  # Styled legend title
    legend.text = element_text(size = 10),  # Styled legend text
    panel.grid.major = element_line(color = "#DDDDDD"),  # Softer grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    panel.background = element_rect(fill = "white", color = NA),  # Panel background
    plot.background = element_rect(fill = "white", color = NA)  # Plot background
  ) +
  geom_text(
    aes(label = Age),  # Add labels showing the number of events
    vjust = -0.5,  # Position labels above the bars
    size = 3,  # Text size
    color = "#333333",  # Text color
    fontface = "bold"  # Bold font
  )
  
#Violin Plot of Age by Cancer Status
ggplot(d, aes(x = as.factor(MCQ220), y = Age_years, fill = as.factor(MCQ220))) +
  geom_violin(trim = FALSE) +
  labs(
    x = "Cancer Status",
    y = "Age",
    title = "Age Distribution by Cancer Status",
    fill = "Cancer Status"
  ) +
  scale_fill_manual(
    values = c("No" = "#FFA500", "Yes" = "#FFDB58"),
    labels = c("Yes" = "Cancer", "No" = "No Cancer")
  ) +
  scale_x_discrete(
    labels = c("Yes" = "Cancer", "No" = "No Cancer")
  ) +
  theme_minimal()
  

  
#Density of Age by Cancer Status
ggplot(d, aes(x = Age_years, fill = as.factor(MCQ220))) +
  geom_density(alpha = 0.5) +
  labs(
    x = "Age",
    y = "Age Density",
    title = "Density of Age by Cancer Status",
    fill = "Cancer Status"
  ) +
  scale_fill_manual(
    values = c("Yes" = "#FFDB58", "No" = "#FFA500"),
    labels = c("Yes" = "Cancer", "No" = "No Cancer")
  ) +
  theme_minimal()
  
#Density of Gender by Cancer Status
ggplot(d, aes(x = as.factor(MCQ220), fill = as.factor(Gender))) +
  geom_bar(position = "fill") +  
  labs(x = "Cancer Status", y = "Proportion", title="Gender Distribution by Cancer Status", fill = "Gender")+
  scale_x_discrete(
  labels = c("Yes" = "Cancer", "No" = "No Cancer")
  )+
  scale_fill_manual(
  values = c("1" = "#FFA500", "2" = "#FFDB58"),
  labels = c("1" = "Male", "2" = "Female")
)+
  theme_minimal()

#Boxplot by Age, Gender, and Cancer Status
ggplot(d, aes(x = as.factor(MCQ220), y = Age_years, fill = as.factor(Gender))) +
  geom_boxplot() +
  labs(
    x = "Cancer Status",
    y = "Age",
    fill = "Gender",
    title = "Age by Gender and Cancer Status"
  ) +
  scale_x_discrete(
    labels = c("Yes" = "Cancer", "No" = "No Cancer")
  ) +
  scale_fill_manual(
    values = c("Male" = "#FFA500", "Female" = "#FFDB58"),
    labels = c("Male" = "Male", "Female" = "Female")
  )+
  theme(plot.title = element_text(hjust = 0.5), legend.position="none", panel.background = element_rect(fill = "white"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())

